Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-10-01 main_new.tex
updates

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for modern marine management and science. Faced with limited
  resources to understand highly processes in the
  water-column, the combination of statistics and autonomous robotics
  provides opportunities for studying aspects related to experimental
  design for effective characterization of oceanographic phenomena. In
  this work we develop methods for efficient spatial sampling applied
  to the mapping of coastal processes by providing informative
  descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the joint excursions of random fields, and
  derive analytical expressions for the expected Bernoulli variance
  reduction in the excursion set of spatial temperature and salinity
  variables. We show how these can be applicable in static designs and
  sequential designs, building on myopic and look-ahead sampling
  strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches followed by results
  from field deployments with an autonomous underwater vehicle as part
  of a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

Traditional data collection at sea has typically been based on static
buoys, Lagrangian floats, or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage but have limited resolution, are limited to
sensing the surface of the ocean, and are impacted by cloud cover. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have advanced the state of sampling and consequently have made
robotics an integral part of ocean observation; our previous work
has contributed to this effort
% using statistical methods in design and data assimilation
\citep{das11b,Das2015,fossum18b,fossuminformation}.
Other statistical work in the oceanographic domain include \cite{wikle2013modern} focusing
on hierarchical statistical models; \cite{sahu2008space}, studying
spatio-temporal models for sea surface temperature and salinity data;
and \cite{mellucci2018oceanic} looking at the statistical prediction
of features using an underwater glider.

The potential that lies in using information-based methods, which we highlight in the current paper, is to do efficient environmental sampling and overcome three common challenges, namely:

\begin{itemize}
\item Full numerical ocean models cannot provide accurate results online if run on robotic sensing platforms, as onboard computers cannot deliver the computational power required, hence statistical proxy models of the environment must be used for doing inference about where to sample.
\item As a priori information about the state of the ocean is uncertain, there is a substantial efficiency gain in reacting to information obtained from measurements taken in-situ. Recently acquired information can be assimilated into statistical models that is used to inform decisions on where to sample sequentially.%\emph{in-situ}; this is usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.
\item For sampling problems related to environmental sensing, the number of choices (i.e. locations, trajectories, and candidate designs) makes the problem combinatorially large, creating a trade-off between optimization (finding the most resource-efficient design to collect necessary data), and computability (arriving at a solution in reasonable time). To successfully resolve features, this trade-off has to be considered in sample design and practice.
\end{itemize}

The combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially-correlated variables and
provide formal measures of uncertainty. In our formulation we use
Gaussian Processes (GPs) in large part because they allow for
efficient implementation and evaluation, in turn enabling
prioritization of sampling efforts which can be formalized into decision-making systems onboard a robotic platform.

@@

Central to understanding the changes taking place in the upper
water-column is an understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography, such as fresh water inflows)
and incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem-related phenomena such as blooms,
anoxic zones and plumes, all of which have significant direct and indirect effects  on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out a lack of sufficient sampling
resolution in time and space as a critical issue for characterizing
ocean phenomena. Even today, policymakers struggle with limited
observability of severe environmental problems such as the recent outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research: https://bit.ly/2KZGJfI} (May 2018), with losses exceeding 12 million
tons of salmon valued at \$80 million (Norwegian Directorate of
Fisheries). This is a case in point of how adequate sampling efforts
could have provided a more accurate assessment and early
warning of changing environmental conditions. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked, and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir1}), where observation is
guided by assimilation of measurements and statistical models.

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network and the sense-plan-act control methodology.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{\ref{fig:envir1} Ocean observation is moving away from single-ship sampling towards more autonomous and collaborative networked operations in order to resolve the numerous bio-geochemical processes and their interactions; this requires autonomy that can reason and make decisions on their own following the sense-plan-act control methodology. \ref{fig:nidelven} The interaction of river and ocean creates processes that are challenging to map, where the combination of robotics and statistics can play a vital role in enabling more effective observation practices.}
\label{fig:envir}
\end{figure}

Our focus in this work is towards spatial characterization of a
frontal system, more specifically, fronts generated by river
plumes. Fig. \ref{fig:nidelven} shows the survey area (the outlet of
the Nidelva river, Trondheim, Norway) where cold freshwater enters
from the river creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coreolis force
\citep{coriolis1835memoire} the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects,
wind, and temperature differences, this boundary often gets
distorted. Prior knowledge about the location and evolution of these
features is therefore highly uncertain, making deterministic planning
challenging.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $5 m^2$ -- $10 km^2$) where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detail at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for
fine resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit the scope to cover only lateral spatial
elements. However, the methodological framework can be extended to
higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a
range of scientific payloads to survey the water column. Typically,
they operate at $\sim 1-3$ m/s, can reach depths from 100--6000 m with
an in-water operation time capacity dependent on survey speed, payload
sensors and mission design. By deploying AUVs together with other
resources, one can augment information obtained from output from synthetic ocean models,
remote-sensing data, or fixed-location sensors on buoys. Traditional
operations and surveys with AUVs are usually limited to observations
along fixed transects on a spatial grid, pre-programmed by the human
operator. By instead using onboard algorithms to continuously
evaluate, update, and refine future sampling locations, one can make
information gathering adaptive and contextual to the survey
requirements \citep{das11b,fossum18b,fossuminformation}. The space of
sampling opportunities is still limited to the spatial grid, but now
there is much more flexibility because of adaptation. For instance, an
AUV could reconstruct or modify a survey line based on what
temperatures it is measuring by using embedded methods in
decision-theoretic planning \citep{py10,Rajan12,Rajan12b}.

@@

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:envir1}). Myopic sampling—i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most—has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward
the designs that improve the classification of water masses as a
means to characterize the frontal zone.

@@

Our goal is to construct adaptive sampling strategies based on continuous evaluation of the EPs. As the EPs provide a description of the boundary uncertainty, the strategies can effectively use this description to prioritize sampling efforts at locations that are ambiguous, making exploration more effective. It is natural to evaluate the expected reduction in the IBV, conditioned on the survey data. The expected updated IBV is
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left( p(\bx;\by)\left( 1-p(\bx;\by)\right) \right) d\bx, 
\end{equation}
where the expectation is over the planned data, and the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey, a criterion for the selection is
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

@@

Calculations of the EPs and the IBV require a model specification. We use GP models, which provides a practical and popular probabilistic approach to modeling environmental data, and allows a closed form
evaluation of the IBVs as derived in Section \ref{sec:sur}. GPs can also be fitted to run onboard robotic vehicles with limited computational assets, such as AUVs.

@@

The Gaussian model is used to represent the bivariate random fields of temperature and salinity. At location $\bx \in \mathcal{M}$, the prior bivariate distribution is
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_T(\bx) \\
    \xi_S(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (i.e. normal) distributed variables
specified by the mean vector and covariance matrix. For the application
of a river plume, the model parameters are specified from a short preliminary survey. The means vary with location, capturing the variability near the river mouth, while the variance parameters and correlation between salinity and temperature are assumed to be constant for all locations. %and within each water mass.

For the spatial dependence among variables, we assume a separable correlation function with the same decay for salinity and temperature,
which is consistent with expectations for  a river plume where the decay is related to the common water masses (Section \ref{sec:case_study}). In the resulting function, we have
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
where for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$, and $\delta_{ij}=0$
otherwise. Moreover, we assume stationary isotropic random fields a priori where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via Euclidean distance $h=\sqrt{||\bx-\bx'||^2}$. With
data and prior knowledge, one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however, that is left for
future work.

We discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area $\Delta$. The
grid is used for the waypoint graph setting the possible design locations. It is also used when solving integral expressions in Eq. (\ref{measVAR}) and
(\ref{sur}), which are then approximated by sums over all grid cells.  Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by the formula: 
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma). %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ contains the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ contains the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

We denote the data at $n_y$ locations by
$\by=(y_{T,1},y_{S,1},\ldots,y_{T,n_y},y_{S,n_y})^t$, representing the
temperature and salinity measurements gathered at the first round of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bG \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bG\bxi)^T\bR^{-1}(\by-\bG\bxi)},
\end{equation}
where $\bG$ is a $2n_y \times 2n$ matrix containing '$1$'s on the
measurement indices of the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_T$ and $r^2_S$ which are measurement error variances of
temperature and salinity observations.

@@

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bG \bmu,\bG \bSigma \bG^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq. (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before inputting the data, the distribution of the conditional mean is
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution \begin{equation}\label{dist_mxi} 
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{T}(\bx) \\
      \mu_{S}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{T}(\bx) & \psi_{T,S}(\bx)\\
      \psi_{T,S}(\bx) & \psi^2_{S}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed-form solution of the design criteria.

@@

Based on Gaussian modeling assumptions, the EPs can be computed at any
location $\bx$. Without additional data, Eq.  (\ref{eq:prob}) gives
\begin{eqnarray}\label{eq:prob_mv0}
 p(\bx) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S) \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx)\sigma_{S}(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx)\sigma_{S}(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
where $\Phi_2$ is the bivariate Gaussian cumulative distribution
function, and the mean and covariance entries are as defined in Eq. (\ref{gp_bar}).

When data $\by$ are available, Eq. (\ref{eq:post_ep}) similarly gives
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by)
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters as defined in Eq. (\ref{gp_upd}) and (\ref{gp_hat}). Here, the probabilities would not only depend on the data but also on the data design defined via matrix $\bG$ in Eq. (\ref{likelihood}). We next derive results that simplify the comparison of several designs.

>>>>

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand dynamic processes in the water-column, the
  combination of statistics and autonomous robotics can provide
  opportunities for studying experimental design for effective
  characterization of oceanographic phenomena. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal processes by providing informative descriptions of spatial
  characteristics of ocean phenomena. Specifically, we define a design
  criterion based on reducing uncertainty in the joint excursions of
  random fields, and derive analytical expressions for the expected
  Bernoulli variance reduction in the excursion set of spatial
  temperature and salinity variables. We show how these can be
  applicable in static and sequential designs, building on myopic and
  look-ahead sampling strategies. We use simulations to study the
  properties and differences between state-of-the-art approaches
  followed by results from field deployments with an autonomous
  underwater vehicle as part of a case study mapping the boundary of a
  river plume. The results demonstrate the ability of the methods to
  effectively inform and execute data-driven environmental sampling.

@@

Traditional data collection at sea has typically been based on static
buoys, Lagrangian floats, or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage but have limited resolution, are limited to
sensing the surface of the ocean and are impacted by cloud cover. The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have
advanced the state of sampling and consequently have made robotics an
integral part of ocean observation; our previous work has contributed
to this effort
% using statistical methods in design and data assimilation
\citep{das11b,Das2015,fossum18b,fossuminformation}.  Other statistical
work in the oceanographic domain include \cite{wikle2013modern}
focusing on hierarchical statistical models; \cite{sahu2008space},
studying spatio-temporal models for sea surface temperature and
salinity data; and \cite{mellucci2018oceanic} looking at the
statistical prediction of features using an underwater glider.

The potential that lies in using information-based methods, which we
highlight in the paper, is to do efficient environmental sampling and
overcome three common challenges, namely:

\begin{itemize}

\item Full numerical ocean models cannot provide accurate results
  online if run on robotic sensing platforms, as onboard computers
  cannot deliver the computational power required, hence statistical
  proxy models of the environment must be used for doing inference
  about where to sample.

\item As a priori information about the state of the ocean is
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken in-situ. Recently
  acquired information can be assimilated into statistical models that
  can be used to inform decisions on where to sample
  sequentially.%\emph{in-situ}; this is usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.

\item For sampling problems related to environmental sensing, the
  number of choices (i.e. locations, trajectories, and candidate
  designs) makes the problem combinatorially large, creating a
  trade-off between optimization (finding the most resource-efficient
  design to collect necessary data), and computability (arriving at a
  solution in reasonable time). To successfully resolve features, this
  trade-off has to be considered in sample design and the practice.

\end{itemize}

The combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially-correlated variables and
provide formal measures of uncertainty. In our formulation we use
Gaussian Processes (GPs) in large part because they allow for
efficient implementation and evaluation onboard a robotic platform.

@@

Central to understanding the changes taking place in the upper
water-column is an understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography, such as fresh water inflows) and
incipient micro-biology driven by planktonic and coastal anthropogenic
input, such as pollution and agricultural runoff. These often result
in a range of ecosystem-related phenomena such as blooms, anoxic zones
and plumes, all of which have significant direct and indirect effects
on society. Walter Munk referred to the 20\textsuperscript{th} century
as the \emph{century of undersampling} \citep{munk2002}, pointing out
a lack of sufficient sampling resolution in time and space as a
critical issue for characterizing ocean phenomena. Even today,
policymakers struggle with limited observability of severe
environmental problems; a prominent example is the recent (May 2018)
outbreak of harmful algal blooms in Northern Norway\footnote{Report
  from the Institute of Marine Research: https://bit.ly/2KZGJfI}, with
losses exceeding 12 million tons of salmon valued at \$80 million
(Norwegian Directorate of Fisheries). This is a case in point of how
adequate sampling efforts could have provided a more accurate
assessment and early warning of changing environmental
conditions. These challenges related to undersampling have been
driving ocean sensing practices towards more autonomous, networked,
and distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir1}), where observation is guided
by assimilation of measurements and statistical models.

\begin{figure}[!b] 
  \centering 
  \subfigure[Illustration of an ocean sensing network and the sense-plan-act control methodology.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{\ref{fig:envir1} Traditional ocean observation based on
    ship-based sampling has recently been augmented by autonomous and
    collaborative networked operations of robotic vehicles to resolve
    numerous bio-geochemical processes and their interactions. The
    sense-plan-act control approach to autonomous control can reason
    and make decisions for efficient onboard adaptive sampling.
    \ref{fig:nidelven} The interaction of river and ocean creates
    processes that are challenging to map, where the combination of
    statistics and robotics can play a vital role in enabling more
    effective observation practices.}
\label{fig:envir}
\end{figure}

Our focus in this work is towards spatial characterization of a
frontal system specifically, fronts generated by river
plumes. Fig. \ref{fig:nidelven} shows the survey area (the outlet of
the Nidelva river, Trondheim, Norway) where cold freshwater enters
from the river, creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coreolis force
\citep{coriolis1835memoire} the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects,
wind, and temperature differences, this boundary often gets
distorted. Prior knowledge about the location and evolution of these
features is therefore highly uncertain, making deterministic planning
challenging.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $5 m^2$ -- $10 km^2$) where
lateral spatial variability tends to dominate. Remote sensing and
synthetic ocean models often find it challenging to provide detail at
this scale \citep{Lermusiaux:2006} and the principal way to resolve
for fine resolution is by direct in-situ observation. At larger
\emph{mesoscale} ($>50 km^2$), both three-dimensional space and time
dynamics are important and can shift substantially, while in the case
of river plumes one can often limit the scope to cover only lateral
spatial elements. However, the methodological framework can be
extended to higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a
range of scientific payloads to survey the water column. Typically,
they operate at $\sim 1-3$ m/s, can reach depths from 100--6000 m with
an in-water operation time capacity dependent on survey speed, payload
sensors and mission design. By deploying AUVs together with other
resources, one can augment information from synthetic ocean models,
remote-sensing data, or fixed-location sensors on buoys. Traditional
operations and surveys with AUVs are usually limited to observations
along fixed transects on a spatial grid, pre-programmed by the human
operator. Instead using onboard algorithms to continuously evaluate,
update, and refine future sampling locations, one can make information
gathering \emph{adaptive} and \emph{contextual} to the survey
requirements \citep{das11b,fossum18b,fossuminformation}. The space of
sampling opportunities is still limited to a grid, but adaptation
allows more flexibility. For instance, an AUV could reconstruct or
modify a survey line based on what temperatures it is measuring by
using embedded methods in decision-theoretic planning
\citep{py10,Rajan12,Rajan12b}.

@@

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:envir1}). Myopic sampling, i.e. step-wise selection of
the path (on the graph) which reduces the information criterion the
most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward the
designs that improve the classification of water masses as a means to
characterize the frontal zone.

@@

Our goal is to construct adaptive sampling strategies based on
continuous evaluation of the EPs. As the EPs provide a description of
the boundary uncertainty, the strategies can effectively use this
description to prioritize sampling efforts at locations that are
ambiguous, making exploration more effective. It is natural to
evaluate the expected reduction in the IBV, conditioned on the survey
data. The expected updated IBV is:

\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left( p(\bx;\by)\left( 1-p(\bx;\by)\right) \right) d\bx, 
\end{equation}
where the expectation is over the planned data, and the probability is
defined in Eq. (\ref{eq:post_ep}).  Considering a set $\mathcal{J}$ of
possible designs for an AUV survey, a criterion for the selection is:
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

@@

Calculations of the EPs and the IBV require a model specification. We
use GP models, which provide a practical and popular probabilistic
approach to modeling environmental data, and allows a closed form
evaluation of the IBVs as derived in Section \ref{sec:sur}. GPs can
also be fitted to run onboard robotic vehicles with limited
computational assets, such as AUVs.

@@

The Gaussian model is used to represent the bivariate random fields of
temperature and salinity. At location $\bx \in \mathcal{M}$, the prior
bivariate distribution is:
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_T(\bx) \\
    \xi_S(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx) \sigma_S(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (i.e. normal) distributed
variables specified by the mean vector and covariance matrix. For the
application of a river plume, the model parameters are specified from
a short preliminary survey. The means vary with location, capturing
the variability near the river mouth, while the variance parameters
and correlation between salinity and temperature are assumed to be
constant for all locations. %and within each water mass.

For the spatial dependence among variables, we assume a separable
correlation function with the same decay for salinity and temperature,
which is consistent with expectations for a river plume where the
decay is related to common water masses (Section
\ref{sec:case_study}). In the resulting function, we have:
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
where for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$, and
$\delta_{ij}=0$ otherwise. Moreover, we assume stationary isotropic
random fields a priori where the correlation function $\rho(\bx,\bx')$
solely depends on $\bx$ and $\bx'$ via Euclidean distance
$h=\sqrt{||\bx-\bx'||^2}$. With data and prior knowledge, one could
possibly fit and estimate parameters of non-stationary or
non-isotropic covariance functions and more complex multivariate
spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however, that is left for
future work.

We discretize the spatial domain $\mathcal{M}$ to a set of $n$ grid
locations $\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell
has area $\Delta$; the grid is used for the waypoint graph setting the
possible design locations. It is also used when solving integral
expressions in Eq. (\ref{measVAR}) and (\ref{sur}), which are then
approximated by sums over all grid cells.  Using vector notation, we
denote the Gaussian temperature and salinity variables on the grid and
its density function by the formula:
\begin{equation}\label{prior}
    \bxi = (\xi_T(\bx_1),\xi_S(\bx_1),\ldots,\xi_T(\bx_n),\xi_S(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma). %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ contains the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ matrix $\bSigma$ contains the
variance-covariances of temperature and salinity variables at all grid
locations.

@@

We denote the data at $n_y$ locations by
$\by=(y_{T,1},y_{S,1},\ldots,y_{T,n_y},y_{S,n_y})^t$, representing the
temperature and salinity measurements gathered at the first round of
nodes in the grid. The conditional model for the data, given the true temperature
and salinity, is described by: 
\begin{equation}\label{likelihood}
\by | \bxi \sim N( \bG \bxi, \bR), %\pi(\by | \bxi)= \frac{1}{(2\pi)^{m}|\bR|^{\frac{1}{2}}}e^{-\frac{1}{2}(\by -\bG\bxi)^T\bR^{-1}(\by-\bG\bxi)},
\end{equation}
where $\bG$ is a $2n_y \times 2n$ matrix containing '$1$'s on the
measurement indices of the grid and '$0$' otherwise. The size
$2n_y \times 2n_y$ covariance matrix $\bR$ contains diagonal entries
$r^2_T$ and $r^2_S$ which are measurement error variances of
temperature and salinity observations.

@@

% The conditional mean $\bm$ will be important in the following
% derivation in Section \ref{sec:sur}.
The calculation of the design criterion in Eq. (\ref{sur}) requires
the marginal distribution of the data which is defined over $\bxi$ to
be $\by \sim N( \bG \bmu,\bG \bSigma \bG^t+\bR)$.  However, it turns
out that a simplified form of the calculation is possible because the
conditional mean in Eq. (\ref{gp_hat}) is a linear (affine) function of
the data $\by$ and the covariance is not a function of the data.
Before inputting the data, the distribution of the conditional mean is:
\begin{equation}\label{distxi} \bm \sim N(\bmu , \bPsi), \hspace{3mm}
  \bPsi=\bSigma \bG^t (\bG \bSigma \bG^t+\bR)^{-1} \bG \bSigma.
\end{equation} 
When considering only grid location $\bx$, we can
extract elements from its mean vector and covariance matrix to get the
bivariate distribution:
\begin{equation}\label{dist_mxi}
  \bm(\bx) \sim N \left( \begin{bmatrix}
      \mu_{T}(\bx) \\
      \mu_{S}(\bx) \end{bmatrix}, \begin{bmatrix}
      \psi^2_{T}(\bx) & \psi_{T,S}(\bx)\\
      \psi_{T,S}(\bx) & \psi^2_{S}(\bx) \end{bmatrix} \right),
\end{equation}
which is important for the derivation of the closed-form solution of the design criteria.

@@

Based on Gaussian modeling assumptions, the EPs can be computed at any
location $\bx$. Without additional data, Eq.  (\ref{eq:prob}) gives:
\begin{eqnarray}\label{eq:prob_mv0}
 p(\bx) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S) \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} \mu_{T}(\bx)\\
\mu_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{T}^2(\bx) & \sigma_{T}(\bx)\sigma_{S}(\bx) \gamma(\bx)  \\
\sigma_{T}(\bx)\sigma_{S}(\bx) \gamma(\bx)  & \sigma_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
where $\Phi_2$ is the bivariate Gaussian cumulative distribution
function, and the mean and covariance entries are as defined in
Eq. (\ref{gp_bar}).

When data $\by$ are available, Eq. (\ref{eq:post_ep}) similarly gives:
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by)
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters as defined in Eq. (\ref{gp_upd}) and
(\ref{gp_hat}). Here, the probabilities would not only depend on the
data but also on the data design defined via matrix $\bG$ in
Eq. (\ref{likelihood}). We next derive results that simplify the
comparison of several designs.
